Regional-scale Base-flow Index of Gauged and Ungauged Basins in Drylands | Regional Base-Flow Index in Arid Landscapes Using Machine Learning and Instrumented Records | Estimating Base-Flow Index in Gauged and Ungauged Dryland Basins

### 2024-12-16 MANUSCRIPT.qmd
finishing up

---
title: Regional-scale Base-flow Index of Gauged and Ungauged Basins in Drylands
author:
  - name: Caelum Mroczek
    orcid: 0009-0005-6695-0134
    corresponding: true
    email: csm428@nau.edu
    roles:
      - Conceptualization
      - Investigation
      - Methodology
      - Formal analysis
      - Visualization
      - Software
      - Data curation
      - Writing - Original Draft
    affiliations:
      - School of Earth and Sustainability, Northern Arizona University, Flagstaff, AZ, USA

  - name: Abraham Springer
    orcid: 0000-0003-4826-9124
    corresponding: false
    roles:
      - Conceptualization
      - Supervision
      - Project administration
      - Funding acquisition
    affiliations:
      - School of Earth and Sustainability, Northern Arizona University, Flagstaff, AZ, USA

  - name: Benjamin Lucas
    orcid: 0000-0002-2021-3076
    corresponding: false
    roles:
      - Software
      - Supervision
    affiliations:
      - Department of Mathematics and Statistics, Northern Arizona University, Flagstaff, AZ, USA

@@

abstract: |
  LOREM IPSUM ######
plain-language-summary: |
  LOREM IPSUM ######

key-points:
  - Base flow sustains dryland river ecology, but assessing its groundwater contribution faces data limitations
  - Recursive filtering and machine learning predict BFI 
  -

@@

Dryland regions, encompassing arid, semi-arid, hyper-arid, and dry sub-humid systems, account for 40% of the Earth’s land surface. These regions are home to approximately 2 billion people globally and constitute the largest terrestrial biome [@iucn-drylands-2019]. Despite supporting diverse ecosystems and human populations, dryland regions face mounting challenges exacerbated by increasing urbanization, expanding agricultural activities, and climate-induced amplification of precipitation patterns [@taylor2013]. This water scarcity is intensifying due to the compounding effects of climate variability and increased groundwater extraction [@taylor2013]. In drylands, groundwater serves as a vital resource for sustaining ecosystems and meeting human needs [@scanlon2006; @yao2018].

@@

To address the challenge of quantifying base flow in ungauged catchments, numerous studies have applied regression and machine learning methods. [@ahiablame2013] found that using a regression model to estimate annual base flow of ungauged catchments was reasonably easy and accurate. @beck2013 overcame the nonlinearity of basin characteristics and improved results of multivariate analyses by using artificial neural networks to estimate BFI globally. @singh2018 implemented a random forest algorithm to predict long-term BFI for ungauged catchments across New Zealand. The application of machine learning in past studies highlights its versatility and effectiveness in addressing ecohydrologic dynamics.

This study develops a technique for estimating BFI in ungauged basins in Arizona and determines the state’s long-term mean BFI. Using a machine learning model trained with the hydrogeologic characteristics of surface water basins, we estimate mean BFI in ungauged basins over a period of record from 1991-2020. Regional patterns and trends in base flow and BFI of instrumented sites are subsequently investigated. These results provide novel analyses of low flow processes in Arizona, furthering our understanding of climate controls on consistent flows in dryland basins.

@@

The state of Arizona, located in the southwest United States, covers a total area of 295,253 km$^2$. Arizona is divided into two primary physiographic provinces: the Colorado Plateau in the northeast, and the Basin-and-Range region in the west and south. The Central Highlands is a transition zone consisting of scattered basins separated by the mountainous foothills of the Mogollon Rim. The Colorado Plateau is dominated by high-elevation desert with an average elevation of 1,936 masl (6,352 ft). Mean temperatures range from -6$^\circ$C (20$^\circ$F) to 26$^\circ$C (80$^\circ$F) and it averages 580 mm (23 in) of precipitation. The Basin-and-Range region has a semi-arid to arid climate with an average elevation of 490 masl (1600 ft). Average temperature ranges from 15$^\circ$C (60$^\circ$F) to 43$^\circ$C (110$^\circ$F) and the region averages 200 mm (8 in) of precipitation annually. ###CITATION###

@@

@@

Annual precipitation and temperature data came from the PRISM climate group at Oregon State University at a resolution of 4 km (<https://prism.oregonstate.edu;> [@daly2008]). The PRISM dataset provides valuable insights into regional climate in ungauged regions and has been shown to perform well across the southwestern US [@buban_PRISM]**.** Instead of the water year, PRISM data uses a calendar-year format, which was adopted for consistency in the water balance. Although this may introduce challenges in the annual estimates due to inter-annual snow storage, the use of long-term annual averages is likely to reduce any potential errors [@reitz2017].

Annual evapotranspiration (ET) data came from TerraClimate, a 4-km grid climatalogical data set [@abatzoglou2018]. TerraClimate uses a Penman-Monteith approach to generate a reference potential evapotranspiration (PET). The PET values were calculated assuming a reference grass surface across the landscape with unlimited water. In the drylands of the southwestern US, PET typically exceeds precipitation annually [@zomer2022].

@@

The implementation of machine learning models to predict hydrologic indices has been successful in past studies [@singh2018; @schmidt_ML_2020; @Rozos2021Machine]. In this work, we used the eXtreme Gradient Boosting (XGBoost) algorithm [@chen2016xgboost] to predict BFI at ungauged locations using catchment characteristics as predictors. The XGBoost algorithm is a decision tree-based ensemble algorithm, which can be adapted for either regression or classification problems. Its efficiency, scalability, and robustness have made it increasingly popular in recent years, with successful applications in environmental modeling tasks such as streamflow forecasting [@szczepanek2022daily; @ni2020streamflow] and land use/land cover classification [@Georganos2018Very].

@@

Machine learning models are prone to overfitting, especially when provided with a large set of predictive features (CITATION). Overfitting can degrade model performance on unseen data and increase the demand for computational resources and memory storage [@feat_selec2017]. Dimensionality reduction offers a robust solution to these challenges and generally falls into two broad categories: feature extraction and feature selection.

@@

The ten most important features (\@suppfig-shap_values) were selected based on their SHAP values and used to train a subsequent model. This refined model demonstrated improved performance and reduced computational time compared to the initial model. The final model, trained on this optimized feature subset, was ultimately used for the analysis presented here.

@@

The long-term average BFI for the 205 gauged reaches across Arizona is illustrated in @fig-instrumented-bfi . The highest BFI values (\>0.9) are found along the Grand Canyon in northwestern Arizona. The highly karstic geology of this region facilitates the rapid movement of subsurface flow to surface water and spring outlets [@chambless_deep-karst_2023]. Relatively high BFI values (\>0.8) are found at the spring-fed headwaters of the Verde River (Del Rio Spring) and the spring-fed headwaters of Fossil Creek. These results are consistent with interpolated BFI values reported by @wolock2003.

The stream reaches of the Little Colorado River Basin (northeastern Arizona) indicate consisently low BFI values (\< 0.2). This is likely due to low-yielding perched aquifers underlying the Defiance Uplift in northeastern Arizona, which are hydrologically connected to surface streams, while the high-yield, confined regional aquifer is much deeper [@blanchard02]. A notable trend emerges along most major rivers in the study area: upstream reaches tend to exhibit higher BFI values, while downstream reaches display lower values. This pattern is presumed to result from greater groundwater-surface water interactions at stream headwaters, often influenced by spring outlets, and the dilution of base flow as water moves downstream (**CITATION**). This trend is particularly evident along the Gila River, Verde River, and Little Colorado River in Arizona.

@@

#### Spatial Trends

TALK ABOUT THE SPATIAL TRENDS IN THE MAP

@@

| Classification Group                    | n   | n_pos | n_neg | perc_pos | perc_neg |
|-------------------------|----------|----------|----------|----------|----------|
| Precipitation - Monsoon Dominated       | 87  | 8     | 21    | 0.092    | 0.241    |
| Precipitation - Snowmelt Dominated      | 118 | 9     | 12    | 0.076    | 0.102    |
| Physiographic Region - Basin and Range  | 156 | 14    | 26    | 0.090    | 0.167    |
| Physiographic Region - Colorado Plateau | 49  | 3     | 7     | 0.061    | 0.143    |
| Climate - Warm-Wet                      | 31  | 2     | 6     | 0.065    | 0.194    |
| Climate - Warm-Dry                      | 55  | 6     | 11    | 0.109    | 0.200    |
| Climate - Cool-Wet                      | 74  | 4     | 9     | 0.054    | 0.122    |
| Climate - Cool-Dry                      | 45  | 5     | 7     | 0.111    | 0.156    |
| Slope - High                            | 102 | 10    | 12    | 0.098    | 0.118    |
| Slope - Low                             | 103 | 7     | 21    | 0.068    | 0.204    |

@@

Given the variations in the period of record across the instrumented network (see SUPPLEMENTAL - Period of Record figure), we analyzed trends in baseflow and BFI in relation to coincident trends in climate variables. Trends are classified as coincident when the direction of the climate variable trend (positive or negative) aligns with the trend observed in baseflow or BFI (@tbl-coincident-trends). This analysis includes both significant and non-significant trends, which is appropriate in environmental studies where the influence of complex, interconnected processes may not always manifest as statistically significant patterns over limited observational periods.

@@

|           | **Climate Variable** | **Coincidence Percentage (%)** |
|-----------|:--------------------:|:------------------------------:|
| BFI       |         AET          |             44.88              |
|           |    Precipitation     |             53.17              |
|           |     Temperature      |             47.32              |
| Base Flow |         AET          |             55.12              |
|           |    Precipitation     |             64.88              |
|           |     Temperature      |             39.02              |

@@

| Classification Group         | n    | R2    | MSE   | RMSE  | MAE   | NSE   | pbias |
|------------------------------|------|-------|-------|-------|-------|-------|-------|
| Climate - Monsoon Dominated  | 3039 | 0.633 | 0.016 | 0.126 | 0.074 | 0.619 | -13.7 |
| Climate - Snowmelt Dominated | 4685 | 0.733 | 0.015 | 0.121 | 0.087 | 0.725 | -3.5  |
| PhysRegion - Basin&Range     | 6147 | 0.733 | 0.016 | 0.127 | 0.084 | 0.724 | -6.3  |
| PhysRegion - CO Plateau      | 1577 | 0.846 | 0.011 | 0.104 | 0.073 | 0.843 | -3.8  |
| Climate - Warm-Wet           | 1506 | 0.693 | 0.014 | 0.117 | 0.077 | 0.685 | -8.2  |
| Climate - Warm-Dry           | 2351 | 0.693 | 0.022 | 0.147 | 0.092 | 0.675 | -11.9 |
| Climate - Cool-Wet           | 2350 | 0.738 | 0.011 | 0.106 | 0.078 | 0.736 | -1.7  |
| Climate - Cool-Dry           | 1517 | 0.831 | 0.012 | 0.111 | 0.078 | 0.827 | -4.3  |
| Slope - High                 | 3795 | 0.776 | 0.012 | 0.111 | 0.079 | 0.771 | -3.3  |
| Slope - Low                  | 3929 | 0.724 | 0.018 | 0.133 | 0.085 | 0.713 | -9.1  |

@@

The predictors used to estimate BFI at ungauged sites were evaluated for their importance in the final XGBoost model, as illustrated in @suppfig-shap_values. The most influential feature for predicting long-term BFI is basin elevation. While elevation itself does not directly affect base flow characteristics, it has consistently been identified as a key predictor in previous BFI studies [@singh2018; @beck2013]. The significance of elevation likely stems from its cumulative effects, such as lower temperatures, increased orographic precipitation, and prolonged snowpack persistence at higher altitudes. Seasonal snowpack duration, in particular, has been shown to strongly correlate with springflow and groundwater recharge in this region [@donovan_karst_2022]. This relationship is further supported by hydroclimate features. Higher temperatures tend to negatively influence BFI, while precipitation exhibits a mixed influence. In some cases, higher precipitation values correlate with lower BFI, likely due to a larger proportion of precipitation contributing to runoff rather than infiltration and groundwater recharge.

@@

#### Predicted Long-term Mean BFI

The long-term BFI (1991-2020) for all stream reaches in Arizona of Strahler stream order 3 and greater are shown in @fig-bfi-streams and the regionalized (HUC-8) long-term BFI is shown in @fig-bfi-huc .

@@

## Conclusion {#sec-conclusion}

@@

## Supporting Information

![SHAP value plot of features used in final model. LC indicates land cover, and soil types are defined by hydrologic soil group.](images/supplemental/shap_summary.png){#suppfig-shap_values}

>>>>

---
title: Regional-scale Base-flow Index of Gauged and Ungauged Basins in Drylands | Regional Base-Flow Index in Arid Landscapes Using Machine Learning and Instrumented Records | Estimating Base-Flow Index in Gauged and Ungauged Dryland Basins
author:
  - name: Caelum Mroczek
    orcid: 0009-0005-6695-0134
    corresponding: true
    email: csm428@nau.edu
    roles:
      - Conceptualization
      - Investigation
      - Methodology
      - Formal analysis
      - Visualization
      - Software
      - Data curation
      - Writing - Original Draft
    affiliations:
      - 'School of Earth and Sustainability, Northern Arizona University, Flagstaff, AZ, USA'

  - name: Abraham Springer
    orcid: 0000-0003-4826-9124
    corresponding: false
    roles: []
    affiliations:
      - 'School of Earth and Sustainability, Northern Arizona University, Flagstaff, AZ, USA'

  - name: Neha Gupta
    orcid: 0000-0002-2247-3509
    corresponding: false
    roles: []
    affiliations:
      - 'Department of Hydrology and Atmospheric Sciences, University of Arizona, Tucson, AZ, USA'

  - name: Benjamin Lucas
    orcid: 0000-0002-2021-3076
    corresponding: false
    roles: []
    affiliations:
      - 'Department of Mathematics and Statistics, Northern Arizona University, Flagstaff, AZ, USA'

@@

abstract: |
  Base flow, sustained by groundwater discharge, is a vital component of river ecosystems, particularly in drylands, where water resources are limited. This study analyzes the instrumented streamflow record in Arizona to assess long-term base-flow index (BFI) trends across gauged catchments. Results indicate that approximately 32% of Arizona’s streamflow originates from groundwater discharge, with significant spatial variability driven by landscape and climatic factors. Base flow  relationships are analyzed with coincident trends in climate variables such as precipitation, evapotranspiration, and temperature. Spatial and climatic trends reveal  variability in base flow contributions, providing insight into groundwater-surface water interactions in arid and semi-arid landscapes. Building on this analysis, we applied machine learning methods to predict BFI in ungauged basins, addressing the challenges of Arizona’s sparse streamgage network. Using the eXtreme Gradient Boosting (XGBoost) algorithm trained on hydroclimate and physiographic predictors, we estimated long-term BFI from 1991 to 2020. This combined approach integrates observational data with predictive modeling to enhance our understanding of base flow processes and provide a framework for water resource management in data-limited regions.

plain-language-summary: |
  Rivers in drylands, such as Arizona, depend on groundwater contributions to maintain flow during dry periods. This portion of streamflow, called base flow, plays a critical role in supporting ecosystems and water availability in these regions. This study examines long-term patterns of base flow in Arizona using data from streamflow monitoring stations. We found that about 32% of Arizona's river flow comes from groundwater, though this varies significantly across the state depending on local climate and landscape features. We also analyzed how base flow is influenced by changes in precipitation, evapotranspiration, and temperature, providing insights into how groundwater and surface water interact in these complex environments. To estimate base flow in basins without monitoring stations, we used machine learning techniques. By training a model on data from monitored sites and corresponding hydroclimate data, we predicted long-term base flow for ungauged areas across Arizona. This innovative approach addresses the challenge of Arizona's sparse monitoring network and provides a valuable tool for understanding and managing water resources in regions with limited data.

key-points:
  - Approximately 32% of Arizona's streamflow originates from groundwater discharge, with significant spatial and climatic variability.
  - A machine learning model (XGBoost) successfully predicted long-term BFI in ungauged basins, addressing gaps in Arizonas sparse streamgage network. 
  - Base-flow trends align most closely with precipitation changes, emphasizing the critical role of climate in shaping groundwater-surface water dynamics in dryland regions.

@@

Dryland regions, encompassing arid, semi-arid, hyper-arid, and dry sub-humid systems, account for 40% of the Earth’s land surface. These regions are home to approximately 2 billion people globally and constitute the largest terrestrial biome [@iucn-drylands-2019]. Despite supporting diverse ecosystems and human populations, dryland regions face mounting hydrologic challenges exacerbated by increasing urbanization, expanding agricultural activities, and climate-induced amplification of precipitation patterns [@taylor2013]. This water scarcity is intensifying due to the compounding effects of climate variability and increased groundwater extraction [@taylor2013]. In drylands, groundwater serves as a vital resource for sustaining ecosystems and meeting human needs [@scanlon2006; @yao2018].

@@

Advancements in machine learning provide tools to predict hydrologic indices in ungauged basins, addressing the limitations of sparse streamgage networks. To tackle the challenge of quantifying base flow in ungauged catchments, numerous studies have applied both regression and machine learning methods. [@ahiablame2013] found that using a regression model to estimate annual base flow of ungauged catchments was reasonably easy and accurate. @beck2013 overcame the nonlinearity of basin characteristics and improved results of multivariate analyses by using artificial neural networks to estimate BFI globally. @singh2018 implemented a random forest algorithm to predict long-term BFI for ungauged catchments across New Zealand. These applications demonstrate the versatility and effectiveness of machine learning in capturing complex ecohydrologic dynamics and improving our understanding of groundwater contributions to streamflow.

This study develops a technique for estimating BFI in ungauged basins across Arizona and evaluates the state’s long-term BFI. Regional trends in baseflow and BFI at instrumented sites are also analyzed, with these trends being linked to coincident trends in precipitation, evapotranspiration (ET~O~​), and temperature over the same periods. Using a machine learning model trained on the hydrogeologic characteristics of surface water basins, we estimate the mean BFI for ungauged basins from 1991 to 2020. This approach helps address the spatial gaps in the streamgage network, which is relatively sparse across the state. The results offer novel insights into low-flow processes in both gauged and ungauged basins, enhancing our understanding of climate controls on consistent flows in Arizona. This study contributes to a more comprehensive view of hydrological dynamics in the context of arid and semi-arid landscapes.

@@

The state of Arizona, located in the southwest United States, covers a total area of 295,253 km$^2$. Arizona is divided into two primary physiographic provinces: the Colorado Plateau in the northeast, and the Basin-and-Range region in the west and south. The Central Highlands is a transition zone consisting of scattered basins separated by the mountainous foothills of the Mogollon Rim. The Colorado Plateau is dominated by high-elevation desert with an average elevation of 1,936 masl (6,352 ft). Mean temperatures range from -6$^\circ$C (20$^\circ$F) to 26$^\circ$C (80$^\circ$F) and it averages 580 mm (23 in) of precipitation. The Basin-and-Range region has a semi-arid to arid climate with an average elevation of 490 masl (1600 ft). Average temperature ranges from 15$^\circ$C (60$^\circ$F) to 43$^\circ$C (110$^\circ$F) and the region averages 200 mm (8 in) of precipitation annually [@az_climate2024].

@@

While our data selection ensures a robust record for this analysis, it also highlights gaps in spatial coverage that the machine learning model aims to address. Arizona has 184 active USGS streamflow stations (as of 2024) across an area of 295,253 km², whereas Indiana has 189 active stations within an area of 94,326 km². This translates to a streamgage density of approximately 2.004 gages per 1,000 km² in Indiana, over three times higher than Arizona's density of 0.623 gages per 1,000 km². This stark difference underscores the relative sparsity of Arizona's streamgage network, particularly in the context of its larger geographic area and the unique hydrological challenges posed by its arid and semi-arid landscapes.

@@

Annual precipitation and temperature data came from the PRISM climate group at Oregon State University at a resolution of 4 km (<https://prism.oregonstate.edu;> [@daly2008]). The PRISM dataset provides valuable insights into regional climate in ungauged regions and has been shown to perform well across the southwestern US [@buban_PRISM]. Instead of the water year, PRISM data uses a calendar-year format, which was adopted for consistency in the water balance. Although this may introduce challenges in the annual estimates due to inter-annual snow storage, the use of long-term annual averages is likely to reduce any potential errors [@reitz2017].

Annual reference evapotranspiration (ET~O~) data came from TerraClimate, a 4-km grid climatalogical data set [@abatzoglou2018]. TerraClimate uses a Penman-Monteith approach to generate a reference evapotranspiration. The ET~O~ values were calculated assuming a reference grass surface across the landscape with unlimited water. In the drylands of the southwestern US, ET~O~ typically exceeds precipitation annually [@zomer2022].

@@

The implementation of machine learning models to predict hydrologic indices has been successful in past studies [@singh2018; @schmidt_ML_2020; @Rozos2021Machine]. In this work, we used the eXtreme Gradient Boosting (XGBoost) algorithm [@chen2016xgboost] to predict BFI at ungauged locations using catchment characteristics as predictors. The XGBoost algorithm is a decision tree-based ensemble algorithm, which can be adapted for either regression or classification problems. This algorithm iteratively builds an ensemble of decision trees, where each tree corrects errors from previous trees to improve predictions [@chen2016xgboost]. Its efficiency, scalability, and robustness have made it increasingly popular in recent years, with successful applications in environmental modeling tasks such as streamflow forecasting [@szczepanek2022daily; @ni2020streamflow] and land use/land cover classification [@Georganos2018Very].

@@

Machine learning models are prone to overfitting, especially when provided with a large set of predictive features [@Ying_2019]. Overfitting can degrade model performance on unseen data and increase the demand for computational resources and memory storage [@feat_selec2017]. Dimensionality reduction offers a robust solution to these challenges and generally falls into two broad categories: feature extraction and feature selection.

@@

SHAP values is a method to explain the prediction of an individual instance by calculating the contribution of each feature to that prediction. The method is based on coalition game theory and is discussed further in @SHAP_values. Here, SHAP values are used for global interpretation of feature importance and feature effects on the model. Global feature importance is produced by the absolute Shapley values of each feature across the dataset, providing a list of features in order of most to least important. Feature effects provide an indication of the relationship between the value of a predicting feature and its impact on the prediction.

The ten most important features [@fig-shap_values] were selected based on their SHAP values and used to train a subsequent model [@tbl-predictors]. This refined model demonstrated improved performance and reduced computational time compared to the initial model. The final model, trained on this optimized feature subset, was ultimately used for the analysis presented here.

@@

The long-term BFI for the 205 gauged reaches across Arizona is illustrated in @fig-instrumented-bfi . The long-term mean BFI is 0.32, indicating that \~32% of long-term streamflow in Arizona likely originates from groundwater discharge and other delayed sources. The highest BFI values (\>0.9) are found along the Grand Canyon in northwestern Arizona. The highly karstic geology of this region facilitates the rapid movement of subsurface flow to surface water and spring outlets [@chambless_deep-karst_2023]. Relatively high BFI values (\>0.8) are found at the spring-fed headwaters of the Verde River (Del Rio Spring) and the spring-fed headwaters of Fossil Creek. These results are consistent with interpolated BFI values reported by @wolock2003.

The stream reaches of the Little Colorado River Basin (northeastern Arizona) indicate consistently low BFI values (\< 0.2). This is likely due to low-yielding perched aquifers underlying the Defiance Uplift in northeastern Arizona, which are hydrologically connected to surface streams, while the high-yield, confined regional aquifer is much deeper [@blanchard02]. A notable tendency emerges along most major rivers in the study area: upstream reaches tend to exhibit higher BFI values, while downstream reaches display lower values. This pattern is presumed to result from greater groundwater-surface water interactions at stream headwaters, influenced by spring outlets, and the dilution of base flow as water moves downstream. This trend is particularly evident along the Gila River, Verde River, and Little Colorado River in Arizona.

@@

@fig-instrumented-trend illustrates the spatial variation in BFI trends across the study area. Statistically significant decreasing trends are observed at 16.1% of sites, while increasing trends are found at 8.8% of sites, with no clear regional patterns for either. In the Basin and Range physiographic region, 9% of sites show increasing trends, while 16.7% exhibit decreasing trends. In the Colorado Plateau region, increasing trends occur at 6.1% of sites, and decreasing trends are observed at 14.3% of sites.

@@

| **Classification Group** | **n** | **n_pos** | **n_neg** | **perc_pos** | **perc_neg** |
|------------|------------|------------|------------|------------|------------|
| Precipitation - Monsoon Dominated | 87 | 8 | 21 | 0.092 | 0.241 |
| Precipitation - Snowmelt Dominated | 118 | 9 | 12 | 0.076 | 0.102 |
| Physiographic Region - Basin and Range | 156 | 14 | 26 | 0.090 | 0.167 |
| Physiographic Region - Colorado Plateau | 49 | 3 | 7 | 0.061 | 0.143 |
| Climate - Warm-Wet | 31 | 2 | 6 | 0.065 | 0.194 |
| Climate - Warm-Dry | 55 | 6 | 11 | 0.109 | 0.200 |
| Climate - Cool-Wet | 74 | 4 | 9 | 0.054 | 0.122 |
| Climate - Cool-Dry | 45 | 5 | 7 | 0.111 | 0.156 |
| Slope - High | 102 | 10 | 12 | 0.098 | 0.118 |
| Slope - Low | 103 | 7 | 21 | 0.068 | 0.204 |

@@

Given the variations in the period of record across the instrumented network (see SUPPLEMENTAL - Period of Record figure), we analyzed trends in baseflow and BFI in relation to coincident trends in climate variables. Trends are classified as coincident when the direction of the climate variable trend (positive or negative) aligns with the trend observed in baseflow or BFI (@tbl-coincident-trends). This analysis includes both significant and non-significant trends, which is appropriate where the influence of complex, interconnected processes may not always manifest as statistically significant patterns over limited observational periods [@ficklin2016].

@@

|           | **Climate Variable** | **Coincidence Percentage (%)** |
|-----------|:--------------------:|:------------------------------:|
| BFI       |        ET~O~         |             44.88              |
|           |    Precipitation     |             53.17              |
|           |     Temperature      |             47.32              |
| Base Flow |        ET~O~         |             55.12              |
|           |    Precipitation     |             64.88              |
|           |     Temperature      |             39.02              |

@@

| **Classification Group** | **n** | **R2** | **MSE** | **RMSE** | **MAE** | **NSE** | **pbias** |
|---------|---------|---------|---------|---------|---------|---------|---------|
| Climate - Monsoon Dominated | 3039 | 0.633 | 0.016 | 0.126 | 0.074 | 0.619 | -13.7 |
| Climate - Snowmelt Dominated | 4685 | 0.733 | 0.015 | 0.121 | 0.087 | 0.725 | -3.5 |
| PhysRegion - Basin&Range | 6147 | 0.733 | 0.016 | 0.127 | 0.084 | 0.724 | -6.3 |
| PhysRegion - CO Plateau | 1577 | 0.846 | 0.011 | 0.104 | 0.073 | 0.843 | -3.8 |
| Climate - Warm-Wet | 1506 | 0.693 | 0.014 | 0.117 | 0.077 | 0.685 | -8.2 |
| Climate - Warm-Dry | 2351 | 0.693 | 0.022 | 0.147 | 0.092 | 0.675 | -11.9 |
| Climate - Cool-Wet | 2350 | 0.738 | 0.011 | 0.106 | 0.078 | 0.736 | -1.7 |
| Climate - Cool-Dry | 1517 | 0.831 | 0.012 | 0.111 | 0.078 | 0.827 | -4.3 |
| Slope - High | 3795 | 0.776 | 0.012 | 0.111 | 0.079 | 0.771 | -3.3 |
| Slope - Low | 3929 | 0.724 | 0.018 | 0.133 | 0.085 | 0.713 | -9.1 |

@@

The predictors used to estimate BFI at ungauged sites were evaluated for their importance in the final XGBoost model, as illustrated in @fig-shap_values. The most influential feature for predicting long-term BFI is basin elevation. While elevation itself does not directly affect base flow characteristics, it has consistently been identified as a key predictor in previous BFI studies [@singh2018; @beck2013]. The importance of elevation aligns with findings from [@beck2013], highlighting its role as a proxy for climate variables such as temperature, precipitation, and snowpack duration. Seasonal snowpack duration, in particular, has been shown to strongly correlate with springflow and groundwater recharge in this region [@donovan_karst_2022]. This relationship is further supported by hydroclimate features. Higher temperatures tend to negatively influence BFI, while precipitation exhibits a mixed influence. In some cases, higher precipitation values correlate with lower BFI, likely due to a larger proportion of precipitation contributing to runoff rather than infiltration and groundwater recharge.

![SHAP value plot of features used in final model. LC indicates land cover, and soil types are defined by hydrologic soil group.](images/shap_summary.png){#fig-shap_values}

@@

#### Predicted Long-term BFI

The regionalized (HUC-8) long-term BFI (1991–2020) is shown in \@fig-bfi-huc, while the long-term BFI for stream reaches with a Strahler stream order of 3 or greater is presented in @fig-bfi-streams. Basins with high BFIs, such as those along the Grand Canyon in the northwestern part of the study area, indicate greater surface water and groundwater interaction. Elevated BFI values are also observed along portions of the Mogollon Rim, a heavily forested region with high precipitation that marks the transition between physiographic regions. Additionally, headwater regions of perennial rivers tend to exhibit higher BFI values. In contrast, low BFI values are found in areas like the Defiance Uplift in northeastern Arizona and the arid southern regions of the state.

@@

@fig-bfi-streams indicates the predicted long-term BFI for stream reaches in Arizona of a Strahler stream order of 3 or greater. Stream-level BFI values often correspond to the long-term BFI observed in HUC-8 basins, as geospatial predictor data were aggregated at the basin level. @fig-bfi-streams highlights stream density across the region, revealing areas with higher concentrations of perennial streams, particularly in regions with favorable hydrogeologic conditions, such as spring-fed systems in the Verde River and the Salt River headwaters. Conversely, lower BFI values are observed in arid regions and those with lower stream densities such as parts of the Little Colorado River Basin, where groundwater recharge is limited [@flint_flint07]. While BFI values and their spatial patterns may align with certain terrain features or climatic patterns, it is important to recognize that BFI is driven by complex interactions among multiple factors [@fig-shap_values].

## Summary & Conclusions {#sec-conclusion}

@@
